A Pandemic-Scale Ancestral Recombination Graph for SARS-CoV-2

### 2025-10-08 latex/main.tex
Updates from Overleaf

[TODO we should mention UShER here: they state multiple times in the GitHub issues 
that they don't handle them at all, so I think we can just say this in some nice way.]
Deletions are an important part of SARS-CoV-2 evolution \cite{Jeronimo2023},
but because they are difficult to distinguish from bioinformatics errors 
and their multi-base nature presents challenges
to the Markovian structure of the LS HMM,
we masked out all gap characters as missing data during primary inference. 
To incorporate the most important deletions into the ARG,
we remapped data using post hoc parsimony at 75 sites involved in deletions
with a frequency of $\geq 1\%$ reported by Li et al.\cite{Li2023indels}
during post-processing (STAR Methods).
Although sites are treated independently in this approach,
runs of mutations to the gap character at adjacent sites
correctly identified many known 
high frequency (Table~\ref{tab:major_dels}) and 
recurrent (Table~\ref{tab:rec_dels}) deletions.
For example, the 4 characteristic deletions associated with 
% FIXME: this is just my notation - is this something recognised more 
% widely? If not, fix to something that is?
% According to Isobel, there is no standard notation.
Alpha (11288+9, 21765+6, 21991+3, 28271+1) are correctly
identified as having a single origin [CITATION?]
(Figure~\ref{fig:subgraph_alpha}).
The 21765+6 deletion (Spike 69-70)
which causes S-Gene Target Failure \cite{Walker2021},
emerges two more times at the origins of the BA.1 and BA.4 variants.
Although most highly recurrent deletions are likely errors,
some known truly recurrent deletions are present,
for example, three deletions in S1-NTD (21765+6, 21991+3, and 22194+3)
associated with antibody escape\cite{McCarthy2021}.
See Document\ref{sec:supplement_dels} for further analysis.

@@

% See https://github.com/jeromekelleher/sc2ts-paper/issues/628
Although uncommon relative to single nucleotide mutations,
deletions are an important component of SARS-CoV-2 evolution,
contributing to the emergence of VOCs by enhancing,
e.g., efficiency for cell entry \cite{Cantoni2022} and
potential for antibody escape \cite{McCarthy2021}.
Using the ARG, we identified major deletion events
(which are propagated to many descendants)
as well as recurrent deletions
(which occur on at least two branches) in the ARG.
In Table~\ref{tab:major_dels},
we list the major deletion events in the ARG
which are each inherited by at least 10,000 descendant samples; and
in Table~\ref{tab:rec_dels},
we list the recurrent deletions in the ARG
which are each observed in at least 10,000 (0.4\%) samples
% [FIXME define how we measured this].
(i.e., for a deletion caused by multiple events in the ARG,
the sum of all the samples inheriting the deletion across all the events).

>>>>

[TODO we should mention UShER here: they state multiple times in the GitHub issues 
that they don't handle them at all, so I think we can just say this in some nice way.]
Deletions are an important part of SARS-CoV-2 evolution \cite{Jeronimo2023},
but because they are difficult to distinguish from bioinformatics errors 
and their multi-base nature presents challenges
to the Markovian structure of the LS HMM,
we masked out all gap characters as missing data during primary inference. 
To incorporate the most important deletions into the ARG,
we remapped data using post hoc parsimony at 75 sites involved in deletions
with a frequency of $\geq 1\%$ reported by Li et al.\cite{Li2023indels}
during post-processing (STAR Methods).
Although sites are treated independently in this approach,
runs of mutations to the gap character at adjacent sites
correctly identified many known 
high frequency (Table~\ref{tab:major_dels}) and 
recurrent (Table~\ref{tab:rec_dels}) deletions.
For example, the 4 characteristic deletions
associated with Alpha \cite{RambautAlpha2020}
% FIXME: this is just my notation - is this something recognised more 
% widely? If not, fix to something that is?
% SHZ: According to Isobel, there is no standard notation.
% So, I'm just following what Rambaut did in this post.
(11288-11296, 21765-21770, 21991-21993)
occur at the origin of Alpha in the ARG (Figure~\ref{fig:subgraph_alpha}).
The 6-base 21765-21770 deletion (Spike $\Delta$H69/V70),
which causes S-Gene Target Failure \cite{Walker2021},
emerges two more times at the origins of BA.1 and BA.4 (Table~\ref{tab:major_dels}).
The ARG also capture genuine recurrent deletions,
for example, three deletions in S1-NTD (21765+6, 21991+3, and 22194+3)
associated with antibody escape\cite{McCarthy2021}.
See Document\ref{sec:supplement_dels} for further analysis.

@@

% See https://github.com/jeromekelleher/sc2ts-paper/issues/628
Although uncommon relative to single nucleotide mutations,
deletions are an important component of
SARS-CoV-2 evolution \cite{Jeronimo2023},
contributing to the emergence of VOCs by enhancing,
e.g., efficiency for cell entry \cite{Cantoni2022} and
potential for antibody escape \cite{McCarthy2021}.
Using the ARG, we identified major deletion events
(which are propagated to many descendants)
as well as recurrent deletions
(which occur on at least two branches) in the ARG.
In Table~\ref{tab:major_dels},
we list the major deletion events in the ARG
which are each inherited by at least 10,000 descendant samples; and
in Table~\ref{tab:rec_dels},
we list the recurrent deletions in the ARG
which are each observed in at least 10,000 (0.4\%) samples
% [FIXME define how we measured this].
(i.e., for a deletion caused by multiple events in the ARG,
the total of all the samples inheriting the deletion across those events).
